Analyse gender in music psychology from journal publication data

### 2025-09-19 manuscript/manuscript.qmd
Worked on discussion: mirrored a few aspects from the introduction and added a few more sentences to improve flow.

Past studies have explored the gender distribution could by due to geographical regions. For example, men out-numbered women more so in high-income compared to low-income countries [@ioannidis2023gender]. However, there are also some unexpected trends: while female representation is lowest in countries like Latin America, Caribbean [in physics journals, @son2022scientific] and in South Korea, Japan, and Austria [in medical journals, @bruck2023bibliometric], females were most highly represented in countries like India and Africa [@bruck2023bibliometric; @son2022scientific]. Human development indicators and gender inequality indicators have also revealed systematic inequalities in authorship, though again perhaps not as fully expected. Although female scientific output is lower in countries ranked in lowest quartile for human development and gender equity, female scientific output is also lower in countries very high in gender equality; better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship]. Another recent paper shows that female authorship rates are most balanced in countries with low to intermediate Human Development Index [@sanchez2024analysis]. Overall, while socioeconomic status of certain countries might be a reason for unbalanced female authorship [@ioannidis2023gender], human development might not fully explain such differences [@bruck2023bibliometric; @sanchez2024analysis; @son2022scientific], and further research is required to better understand the trends.

Another aspect that differs between distribution of gender in authorship is the research methodologies and topics. In terms of methodologies, women are are less likely to publish quantitative and experimental studies, but rather publish qualitative work [@ashmos2011gendering; @nunkoo2020three; @sebo2020gender; @thelwall2019gender; @zhang2023gender], perhaps as they demonstrate better relational skill involved in qualitative studies, like prolonged relationships and emotional ties with research participants [@sebo2020gender]. Interestingly, such stereotypes are occur in the opposite direction: when assessing fictive research abstracts-one with quantitative and one with a qualitative design- scientific quality of the qualitative abstract was ranked more trustworthy and accurate when assigned to a female author [@johansson2002gender]. In an analyses of article keywords of medical journals, female-authored articles were more associated with keywords more related to healthcare-related themes like patient involvement, insurance, and quality-of-care compared to males [@bruck2023bibliometric]. Similarly, in computational linguistics, woman are more likely to publish on speech, social, and conversational topics than males; the latter are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. These trends could be explained by the idea that female researchers often value research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. Overall, research shows that women choose qualitative research on social and societal topics, while males author papers using more quantitative methods on theoretical and analytical topics.

Looking more broadly past the information within papers themselves, other research provides some further explanatins for discrepancies in female/male authorships. Despite efforts to give fair credit for author contributions [@ni2021gendered] using, for example, the Contributor Roles Taxonomy (CRediT) system, assigning authorship can be unclear. Women are more likely to experience authorship disagreements, for example, when to decide authorship: women prefer to discuss in earlier stages, while men choose authorship at the final stage [@ni2021gendered]. Another reason for the gender gap could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Additionally, women faculty perform more university, campus, and departmental services than men in general [@guarino2017faculty]. Fortunately, there are strides to improve the current situation, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison]. To improve academic equity issues, it is worthwhile focusing on our discipline and to assess within subdisciplines to provide more 'localised' support to improve the equity in music psychology. Indeed, the past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

Panel A in Figure 3 shows a small number of countries, such as Australia, China, and Canada, with a high odds ratio for female authorship. At the other end of the spectrum, Japan, Belgium, Austria, Sweden, and Turkey all have significantly lower odds ratios for female authors. There is no simple rationale immediately apparent (e.g., country size, membership in a particular political bloc, gender equality index, or domination of articles by a single eminent scholar from one country) that would explain these differences, but this representation provides some starting points for discussion and may highlight the impact of academic career structures and principles implemented in these countries. Panel B shows data aggregated by continent, where Africa has meagre representation (52 authors), Oceania consists only of Australia and New Zealand, and the Americas are represented almost entirely by the US and Canada. While this level of presentation may be too coarse for detailed analysis, it does offer a broad summary of the current situation.

@@

Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented. As such representations differ across different disciplines, it is worthwhile to diagnose the situation in music psychology. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006].

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% [@eerola2026inprep]) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. By contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science].

There were no substantial gender differences in the citation patterns. However, one notable observation is that first-authored articles by women were significantly more likely to be published open access, with an odds ratio of 1.49, a result that warrants further exploration. This effect does not appear to be tied to specific journals. Although the odds ratio for women publishing open access is highest in _Music and Science_, the broader trend suggests a temporal shift: since around 2010, women have increasingly overtaken men in publishing open access. The data does not provide any direct reasons for this, although the legislation in few of the European nations (United Kingdom, EU) require open access publishing and when women have higher odds ratio for being first authors, this might be due to the high number of female from these countries. While past surveys among, for instance, UK academics have not indicated substantial differences in experiences with open access [@zhu2017support], financial and other inequalities in emerging countries have been suggested to limit women's capacity to publish in open access journals [@vuong2021adopting]. But nothing in the extant literature would point out to the reverse pattern, which requires further investigation about the reasons for adopting open access publishing models.

@@

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on care and nurturing [@gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he].

@@

Current findings present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are comparable to psychology in general, musicology (and the closely associated disciplines such as music theory, popular music and ethnomusicology) tend to have less equal division of gender in overall authorship; a recent analysis of the last 25 years of articles in 15 most eminent musicology journals amounting to 3,515 articles suggests that female authors are far more infrequent (34.8% [33.3%-36.3%]) than males (65.2% [63.7%-66.7%]) [@eerola2026inprep]. The present findings should be used in conjuction of ...

>>>>

Past studies have explored the gender distribution could by due to geographical regions. For example, men out-numbered women more so in high-income compared to low-income countries [@ioannidis2023gender]. However, there are also some unexpected trends: while female representation is lowest in countries like Latin America, Caribbean [in physics journals, @son2022scientific] and in South Korea, Japan, and Austria [in medical journals, @bruck2023bibliometric], females were most highly represented in countries like India and Africa [@bruck2023bibliometric; @son2022scientific]. Human development indicators and gender inequality indicators have also revealed systematic inequalities in authorship. women’s academic success is correlated  positively with national indicators of gender equity [@chan2020]. However, this does not seem to be as clearly reflected in authorships. Female scientific output is lower in countries ranked in lowest quartile for human development and gender equity, female scientific output is also lower in countries very high in gender equality; better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship]. Another recent paper shows that female authorship rates are most balanced in countries with low to intermediate Human Development Index [@sanchez2024analysis]. Overall, while socioeconomic status of certain countries might be a reason for unbalanced female authorship [@ioannidis2023gender], human development might not fully explain such differences [@bruck2023bibliometric; @sanchez2024analysis; @son2022scientific], and further research is required to better understand the trends.

Another aspect that differs between distribution of gender in authorship is the research methodologies and topics. In terms of methodologies, women are are less likely to publish quantitative and experimental studies, but rather publish qualitative work [@ashmos2011gendering; @nunkoo2020three; @sebo2020gender; @thelwall2019gender; @zhang2023gender], perhaps as they demonstrate better relational skill involved in qualitative studies, like prolonged relationships and emotional ties with research participants [@sebo2020gender]. Interestingly, such stereotypes occur in the opposite direction: when assessing fictive research abstracts-one with quantitative and one with a qualitative design- scientific quality of the qualitative abstract was ranked more trustworthy and accurate when assigned to a female author [@johansson2002gender]. In an analyses of article keywords of medical journals, female-authored articles were more associated with keywords more related to healthcare-related themes like patient involvement, insurance, and quality-of-care compared to males [@bruck2023bibliometric]. Similarly, in computational linguistics, woman are more likely to publish on speech, social, and conversational topics than males; the latter are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. These trends could be explained by the idea that female researchers often conduct research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. Overall, research shows that women choose qualitative research on social and societal topics, while males author papers using more quantitative methods on theoretical and analytical topics.

Looking more broadly past the information within papers themselves, other research provides some further explanations for discrepancies in female/male authorships. Despite efforts to give fair credit for author contributions [@ni2021gendered] using, for example, the Contributor Roles Taxonomy (CRediT) system, assigning authorship can be unclear. Women are more likely to experience authorship disagreements, for example, when to decide authorship: women prefer to discuss in earlier stages, while men choose authorship at the final stage [@ni2021gendered]. Another reason for the gender gap could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Additionally, women faculty perform more university, campus, and departmental services than men in general [@guarino2017faculty]. Fortunately, there are strides to improve the current situation, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison]. To improve academic equity issues, it is worthwhile focusing on our discipline and to assess within subdisciplines to provide more 'localised' support to improve the equity in music psychology. Indeed, the past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

Panel A in Figure 3 shows a small number of countries, such as Australia, China, and Canada, with a high odds ratio for female authorship. At the other end of the spectrum, Japan, Belgium, Austria, Sweden, and Turkey all have significantly lower odds ratios for female authors. There is no simple rationale immediately apparent (e.g., country size, membership in a particular political bloc, gender equality index, or domination of articles by a single eminent scholar from one country) that would explain these differences, but this representation provides some starting points for discussion and may highlight the impact of academic career structures and principles implemented in these countries. Panel B shows data aggregated by continent, though we note that the representation is markedly different between continents. Africa has meagre representation of only 52 authors, Oceania consists only of Australia and New Zealand, and the Americas are represented almost entirely by the US and Canada. While this level of presentation may be too coarse for detailed analysis, it does offer a broad summary of the current situation. Overall, Oceanie and Americas have higher female author representation, while Asia and Europe have more a male representation.

@@

Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented in aspects such as authorships. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006]. As male-female authorship representations differ across different academic disciplines, it is worthwhile to diagnose the situation in music psychology.

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% [@eerola2026inprep]) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This suggests that the more 'leadership' author position is more likely to go to the male author.

There were no substantial gender differences in the citation patterns. However, one notable observation is that first-authored articles by women were significantly more likely to be published open access, with an odds ratio of 1.49, a result that warrants further exploration. This effect does not appear to be tied to specific journals. Although the odds ratio for women publishing open access is highest in _Music and Science_, the broader trend suggests a temporal shift: since around 2010, women have increasingly overtaken men in publishing open access. The data does not provide any direct reasons for this, although the legislation in few of the European nations (United Kingdom, EU) require open access publishing and when women have higher odds ratio for being first authors, this might be due to the high number of female from these countries. While past surveys among, for instance, UK academics have not indicated substantial differences in experiences with open access [@zhu2017support], financial and other inequalities in emerging countries have been suggested to limit women's capacity to publish in open access journals [@vuong2021adopting]. The idea that publishing in a more open and transparent way benefit society in general [@pervsic2023open], and this fits with the general idea that female researchers value research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. However, nothing in the extant literature would point out to the reverse pattern, which requires further investigation about the reasons for adopting open access publishing models.

@@

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on (health)care and nurturing [@bruck2023bibliometric; @gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. It perhaps might be interesting to note that 'emotion' has an equal balance between men and women. Although this may come as a surprise hen coming from different disciplines, music emotion can be a relatively 'analytical' aspect in music psychology, where emotion from music is attempted to be decoded in a more analytical way (toiviainen papers?), as well as being used in a more 'emotional' sense. this is the same for the keyword 'expression'. Although on first glance, it might fit the stereotype that this keyword might be more related to females, the 'expressivity' is again related to a more analytical coding aspect <cite Wanderley/thompson etc.>

In summary, music psychology has similar groupings of topic for female and male authors as in other disciplines of psychology and medicines. Nonetheless, it should a positive trend. Below we also discuss how this information can help benefit the general field towards more equity with academia.

@@

Current findings present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are comparable to psychology in general, musicology (and the closely associated disciplines such as music theory, popular music and ethnomusicology) tend to have less equal division of gender in overall authorship; a recent analysis of the last 25 years of articles in 15 most eminent musicology journals amounting to 3,515 articles suggests that female authors are far more infrequent (34.8% [33.3%-36.3%]) than males (65.2% [63.7%-66.7%]) [@eerola2026inprep]. The present findings should be used in conjunction of ...
